---
title: "Coding Assistants in RStudio and Visual Studio Code"
author: "Ina Krapp"
format:
  revealjs: 
    theme: simple
    self-contained: true
editor: visual
---

## What is Ollama?

**Ollama:**

::: incremental
-   Offers a ChatGPT-like interface (for Windows)

-   Runs models locally, guaranteeing data privacy

-   Downloads models you want to run automatically

-   Has a wide variety of models: We will use gemma3:1b here

-   **Download to try it out:** [Download Ollama here](https://ollama.com/download "Download Ollama here")
:::

## Model families:

::: incremental
-   American: Gemma (Google), GPT-OSS (OpenAI), Llama (Meta)

-   Training: Predominantly Kenya

-   Chinese: Qwen (Alibaba), Deepseek

-   French: Mistral (Mistral AI)
:::

## Model sizes:

::: incremental
-   Large Language Model's size is given in billion parameters
-   gemma3:1b: Gemma3 model with 1 billion parameters
-   My rule of thumb: For 1B parameters, plan 1GB RAM
-   If you're uncertain, look up RAM under 'About your PC'
-   Ollama fails 'gracefully', but still good to keep in mind
:::

## Which model should I choose?

::: incremental
-   Models run locally: Speed depends on your computer
-   Larger models are usually more accurate, but slower
-   Newer models tend to perform better
:::

::: {.fragment .fade-up}
Less relevant today:
:::

::: incremental
-   Specialized coding vs. general models
-   Language preferences
:::

## RStudio - Chat with a LLM:

First, install the ellmer package and start a chat in it:

```{r, echo = TRUE}
#install.packages("ellmer")
library("ellmer") 
chat <- chat_ollama(model = "gemma3:1b")
chat$chat("How can I create a plot of the diamonds dataset in R? Answer briefly.")
```

## RStudio - let a LLM add code to R files:

To enable a LLM to add code to your R file, you need the gander package:

```{r, echo = TRUE}
#install.packages("gander")  
library(gander)
options(.gander_chat = ellmer::chat_ollama(model = "gemma3:1b"))
```

::: {.fragment .fade-up}
This will also automatically add any data from the environment and code from the file to the prompt.
:::

::: {.fragment .fade-up}
**Let's try it out!**
:::

::: {.fragment .fade-up}
Note: Some models add 'R' when they create R code.
:::

## RStudio - Setting a Keyboard Shortcut

Gander is intended to be used with a keyboard shortcut:

-   In RStudio, Tools \> Modify Keyboard Shortcuts \> Search "gander"

-   Choose a shortcut: The package author recommends Ctrl+Alt+G (or Ctrl+Cmd+G on macOS), I use Ctrl+G

## RStudio - Saving a chat in the R Profile:

-   Gander requires the chat to be restarted each time a new R Session starts.

-   To avoid this, you can save the chat in your R Profile:

```{r, echo = TRUE}
usethis::edit_r_profile()
```

-   Then, copy and paste the chat option into the file that opened. Below again the chat option we use:

```{r}
options(.gander_chat = ellmer::chat_ollama(model = "gemma3:1b"))
```

-   Modify the profile later to change the chosen model

## RStudio - Troubleshooting:

-   You might have to install Rtools from here if you don’t have it already: [Download Rtools here](https://cran.rstudio.com/bin/windows/Rtools/)
-   Updating the glue, rlang and cli packages may be required
-   You can install the development version of gander with:

```{r, echo = TRUE}
#install.packages("pak")
#pak::pak("simonpcouch/gander")
```

## Visual Studio Code:

Large Language Models in Visual Studio Code are available as 'Continue' Extension:

::: incremental
-   Extensions \> Search 'Continue'

-   Select Ollama as Provider
:::

## Visual Studio Code - Using Models:

::: incremental
-   Unlike RStudio, Visual Studio Code uses some default Ollama models: llama 3.1 8b for Chat, qwen2.5-coder:1.5b-base for Code autocompletion

-   Advantage: Requires no explicit Chat starting, runs different models

-   Disadvantage: Easy to get confused
:::

## Visual Studio Code - Changing Models:

::: incremental
-   In a Chat, click default model, then click on ‘add chat model’.

-   Under ‘Provider’, select Ollama if it is not selected already.

-   Choose a model, or click on ‘Autodetect’ to make the models available you tried out in Ollama so far.
:::

::: {.fragment .fade-up}
**Let's try it out!**
:::

## Visual Studio Code - Working with Continue:

::: incremental
-   In 'Chat' mode, click on or write '\@' to provide a file as context to your prompt.

-   Use 'Ctrl+Alt+Space’ to start a code completion.

-   Continue also offers 'Plan' and 'Agent' modes instead of Chat.

-   'Agent' mode can run code - be careful about that!
:::

## What's next?

-   Many new models, like gemma3:4b, are multimodal:

-   They can process images and videos

-   Ollama requires terminal to do that:

```{bash, echo = TRUE, eval=FALSE}
 ollama run gemma3:4b "What's in this image?" "path/to/image.png"
```

-   If you are interested in other agents, you can take a look at Goethe University's [Ai Toollab](https://gki.studiumdigitale.uni-frankfurt.de/ai-toollab/)

## Helpful links:

-   To download R: [Download R here](https://cran.uni-muenster.de/)

-   To download Python: [Download Python here](https://www.python.org/downloads/)

-   To download RStudio: [Download RStudio here](https://posit.co/download/rstudio-desktop/)

-   To download Visual Studio Code: [Download Visual Studio Code here](https://code.visualstudio.com/)

-   To browse models available in Ollama: [Search Ollama models here](https://ollama.com/search)
